Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-08-13 chapters/chapter01.tex
small edits decsription evol process

The last decade has witnessed a particular focus of analyses of viral evolutionary dynamics on integrating concepts of phylogenetic and epidemiological dynamics, a research direction originally coined `phylodynamics'. %PL: cite Grenfell, Science
By aiming at the development a full quantitative understanding of the processes that shape the epidemiology and evolution of RNA viruses,
phylodynamics has now become a burgeoning area of research.
In order to perform comprehensive and quantitative analyses of the interaction between the ecological and evolutionary dynamics of RNA viruses, new computational developments are needed... [PL: get some inspiration from  DOI: 10.1371/journal.pcbi.1002947 to add some meat to this paragraph / finish it]...
% http://www.ploscompbiol.org/article/info%3Adoi%2F10.1371%2Fjournal.pcbi.1002947
In order to position these recent developments, I first provide a general and brief account of the field of molecular evolution and phylogenetics and then introduce recent developments and models specific for rapidly-evolving RNA viruses, and how this culminated in to a comprehensive statistical framework for phylodynamics. Finally, I outline how the current doctoral thesis work extends and complements this framework in the objectives.
The remainder of this chapter provides a more mathematically oriented description on how to model evolution evolution as a stochastic process on trees, the computational aspects involved and how to treat the problem from a Bayesian statistical perspective.

@@

[PL: briefly introduce the field of molecular evolution, touching upon the essence of some basic evolutionary concepts (mutation<->substitution, selection, fitness), and how phylogenetics can be useful int this context. 
Describe the very basics of a tree (branches, nodes, taxa, rooted, outgroup), and provide some history on phylogenies and what they are used for. I would suggest to start by showing a figure combining Darwin's speciation tree in one panel, because that directly relates to the evolutionary and speciation process you need to bring up in the paragraph above, and the first estimated tree, which I believe is from Fitch in seventies science paper.
And then follow up on this with a brief account of phylogenetic inference -- bring on the classics, acknowledge some milestones: distance-based trees, parsimony inference, Joe and Likelihood inference (no formal stuff/math just yet), ...
This is also where you can describe the computational burden and that table.

@@

Molecular phylogenetic modelling typically starts with an assumption that the observed sequence data has been generated by some stochastic process, which emits characters over time.
We can think of stochastic processes as functions of time, which is regarded as a deterministic argument, whose values are random (non-deterministic) variables. 
It is thus a statistical model of a random development, or evolution, in time.
Phylogenetic inference often resorts to processes where time is considered to be continuous and the indexed set of outcomes, called the state space, is countable and finite (discrete).
% The stochastic process $X$ is a collection $\left\{ X(t):\; t\in T\right\} $, where each drawn value $X(t)$ is a random variable drawn from state space $\mathcal{E}$, with $K$ possible values, indexed by orderer time $T$.
With that in mind we can give a formal definition:

@@

In the simplest case, for which the unit of evolution is a single nucleotide, we have $\mathcal{E}=\left\{ A,C,G,T\right\}$ and $K=4$.
The particular sites of the alignment are then considered to be stochastically independent realisations of the process $X$, leading to the observed data (see Figure~\ref{fig:alignment}).

\begin{figure}[H]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
    \xymatrix@R+1pc{ 
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & A \\
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & G \\
\color{gray2}A  \ar[rr]_{ \color{gray2}\begin{array}{c}
A\rightarrow G\\
G\rightarrow C
\end{array}} &&  C
    } % 
\end{displaymath}
}% END: Large
\endgroup
\caption{
{ \footnotesize 
{\bf Evolution at a single site.} Over time charactes at the site change, leading to the observed data (black). Some of the mutations may be silent. Each site evolves independently.
}% END: footnotesize
}
\label{fig:alignment}
\end{figure}

As Figure~\ref{fig:alignment} depicts not all of the differences can be observed, as due to multiple substitutions (multiple hits) some of the changes are silent.
To accurately describe these processes one needs a models of evolution that can account for those multiple hits.

>>>>

The last decade has witnessed a particular focus of analyses of viral evolutionary dynamics on integrating concepts of phylogenetic and epidemiological dynamics, a research direction originally coined `phylodynamics'. %PL: cite Grenfell, Science
By aiming at the development a full quantitative understanding of the processes that shape the epidemiology and evolution of RNA viruses,
phylodynamics has now become a burgeoning area of research.
In order to perform comprehensive and quantitative analyses of the interaction between the ecological and evolutionary dynamics of RNA viruses, new computational developments are needed... [PL: get some inspiration from  DOI: 10.1371/journal.pcbi.1002947 to add some meat to this paragraph / finish it]...
% http://www.ploscompbiol.org/article/info%3Adoi%2F10.1371%2Fjournal.pcbi.1002947
In order to position these recent developments, I first provide a brief general account of the field of molecular evolution and phylogenetics and then introduce recent developments and models specific for rapidly-evolving RNA viruses, and how this culminated in to a state-of-the-art statistical framework for phylodynamics. As a final part of the general introduction, I outline how the current doctoral thesis work extends and complements this framework in the objectives.
The remainder of the chapter provides a more mathematically-oriented description of how to model evolution evolution as a stochastic process on trees, the computational aspects involved, and how to treat the problem from a Bayesian statistical perspective.

@@

[PL: briefly introduce the field of molecular evolution, touching upon the essence of some basic evolutionary concepts (mutation<->substitution, selection, fitness), and how phylogenetics can be useful int this context. 
Describe the very basics of a tree (branches, nodes, taxa, rooted, outgroup), and provide some history on phylogenies and what they are used for. I would suggest to start by showing a figure combining Darwin's speciation tree in one panel, because that directly relates to the evolutionary and speciation process you need to bring up in the paragraph above, and the first estimated tree, which is from Fitch from this science paper:
Fitch WM, Margoliash E. The construction of phylogenetic trees - a generally applicable method utilizing estimates of the mutation distance obtained from cytochrome c sequences. Science 1967;155:279-284.
And then follow up on this with a brief account of phylogenetic inference -- bring on the classics, acknowledge some milestones: distance-based trees, parsimony inference, Joe and Likelihood inference (no formal stuff/math just yet), ...
This is also where you can describe the computational burden and that table.

@@

In order to provide a formal basis for the work in the research chapters of this thesis, this chapter proceeds with a description of general stochastic models of character substitution processes in phylogenies. 
Molecular phylogenetic modeling typically starts with an assumption that the observed sequence data has been generated by some stochastic process, which emits characters over time.
We can think of these stochastic processes as functions of time, which is regarded as a deterministic argument, whose values are random (non-deterministic) variables. 
It is thus a statistical model of a random development, or evolution, in time.
Phylogenetic inference often resorts to processes where time is considered to be continuous and the indexed set of outcomes, called the state space, is countable and finite (discrete).
% The stochastic process $X$ is a collection $\left\{ X(t):\; t\in T\right\} $, where each drawn value $X(t)$ is a random variable drawn from state space $\mathcal{E}$, with $K$ possible values, indexed by orderer time $T$.
With that in mind we can provide a formal definition:

@@

In the simplest case, for which the unit of evolution is a single nucleotide, we have $\mathcal{E}=\left\{ A,C,G,T\right\}$ and $K=4$.
The particular sites of the sequence alignment are then considered to be stochastically independent realizations of the process $X$, leading to the observed data (see Figure~\ref{fig:alignment}).

\begin{figure}[H]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
    \xymatrix@R+1pc{ 
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & A \\
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & G \\
\color{gray2}A  \ar[rr]_{ \color{gray2}\begin{array}{c}
A\rightarrow G\\
G\rightarrow C
\end{array}} &&  C
    } % 
\end{displaymath}
}% END: Large
\endgroup
\caption{
{ \footnotesize 
{\bf Evolution at a single site.} Over time characters at the site change, leading to the observed data (black). Some of the mutations may be silent. Each site evolves independently.
}% END: footnotesize
}
\label{fig:alignment}
\end{figure}

As Figure~\ref{fig:alignment} depicts not all of the differences can be observed, as due to multiple substitutions (multiple hits) some of the changes are silent.
%PL: better not use silent as this also used for 'non synonymous substitutions'
To accurately describe these processes one needs a models of evolution that can account for those multiple hits.
